### 2023-12-13 sections/1-introduction.tex
some edits

Often dubbed ``the Big Data explosion,'' rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$.

@@

In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating that computer architecture will not ``catch up'' to our computational needs in the near future.
As a result, analysis of these large datasets requires better algorithms.

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4~\cite{OpenAI2023GPT4TR}, and LLAMA-2~\cite{Touvron2023Llama2O}, similarity search can become a useful tool for understanding and verifying the validity of model embeddings. 
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

We now define two types of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN). 
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \ \setminus \ S_q$, $f(x, q) > \max\{f(y, q): y \in S_q \}$;
that is, for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
On the other hand, $\rho$-nearest neighbors search aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

Previous works have used the term \emph{approximate} search to refer to $\rho$-nearest neighbor search, but in this paper, 
we reserve the term \emph{approximate} for $k$-NN search algorithms which may miss some of the $k$ nearest elements, and erroneously include some of the elements other than the truly nearest $k$.
In other words, we call a $k$-NN search algorithm \emph{approximate} if there exists a query, dataset, distance function, or value of $k$ for which its set $H$ of returned points satisfies the condition $H \triangle S_q \neq \emptyset$, where $\triangle$ denotes a symmetric difference.
When we refer to an \emph{exact} $k$-NN algorithm, we describe an algorithm which, no matter the query, dataset, distance function, nor value of $k$, returns precisely the $k$ nearest neighbors to that query;
no closer points are elided and no further points are included.

$k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Na\"{i}ve implementations of $k$-NN, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets.
While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not exploit the geometric and topological structure inherent in large datasets.
Approximate search is sufficient for some applications, but the need for efficient, \emph{exact} search remains.
In particular, while approximate $k$-NN may agree with exact $k$-NN in a majority vote for large values of $k$, it may be sensitive to local perturbations with smaller values of $k$, and small values of $k$ may be necessary when classes are not well-separated.
There is some evidence that distance functions which are non-metrics, such as cosine distance, perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance};
this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could exhibit suboptimal accuracy.

@@

This paper focuses on four novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-nearest neighbors search algorithms in CHESS, as well as improved genericity across distance functions.
Additionally, we prove a guarantee about the scaling behavior of cluster radii. 
Finally, we provide a comparison of CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} and HNSW, on several datasets from the ANN benchmarks suite~\cite{aumuller2020ann}.

>>>>

Often dubbed ``the Big Data explosion,'' rapidly advancing technologies have enabled researchers to collect complex data at an unprecedented scale. 
This is especially true in the field of astronomy. 
% Najib: We should mention datasets in this order: neural -network embeddings, biological, radio-ml, and then astronomical.
The Sloan Digital Sky Survey (SDSS)~\cite{blanton2017sdss}, for example, in its APOGEE~\cite{alam2015eleventh} dataset, has collected near-infrared spectra of approximately 430,000 stars in approximately 9,000 dimensions. 
Their MaNGA dataset also contains spectral measurements across the face of each of about 10,000 galaxies across 2,700 deg$^2$.

@@

% Najib: Switch the next paragraph for this:
% In many fields, the sizes of datasets are growing exponentially, and this increase in the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}.
% This indicates that computer architecture will not ``catch up'' to computational needs in the near future.
% Thus, analysis of these large datasets requires better algorithms.
In many fields, the rate of data collection outpaces the rate of computing performance improvements predicted by Moore's Law~\cite{brescia2012extracting}, indicating that computer architecture will not ``catch up'' to our computational needs in the near future.
As a result, analysis of these large datasets requires better algorithms.

Many researchers are especially interested in similarity search on these datasets. 
Similarity search enables a variety of applications, including recommendation and classification systems. 
% Najib: The previous sentence needs a citation.
Additionally, with the rise of large language models such as GPT-3~\cite{2020arXiv200514165B}, GPT-4~\cite{OpenAI2023GPT4TR}, and LLAMA-2~\cite{Touvron2023Llama2O}, similarity search can become a useful tool for understanding and verifying the validity of model embeddings. 
% Najib: As the cardinalities and dimensionalities ...
As the sizes and dimensionalities of datasets have grown, however, efficient and accurate similarity search has become extremely challenging; 
even state-of-the-art algorithms exhibit a steep recall versus throughput tradeoff~\cite{ishaq2019clustered}.

% Najib: Rework this paragraph. I don' want any mathematically notation in the introduction, particularly because this is repeated in Methods, The Search Problem.
We now define two types of similarity search: $k$-nearest neighbor search ($k$-NN) and $\rho$-nearest neighbor search ($\rho$-NN). 
Given a query $q$, along with a distance function $f$ defined on a dataset $\textbf{X}$, $k$-nearest neighbors search aims to find 
the set $S_q$ such that  $|S_q| = k$ and $\forall x \in \textbf{X} \ \setminus \ S_q$, $f(x, q) > \max\{f(y, q): y \in S_q \}$;
that is, for a given $k$, find the $k$ closest points to $q$ in $ \textbf{X}$.
On the other hand, $\rho$-nearest neighbors search aims to find the set $\{x \in \textbf{X}: f(q, x) \leq \rho \}$;
that is, find all points in $\textbf{X}$ that are at most a distance $\rho$ from $q$.

Previous works have used the term \emph{approximate} search to refer to $\rho$-nearest neighbor search, but in this paper, 
% Najib: we reserve the term \emph{approximate} for search algorithms which do not exhibit perfect recall.
we reserve the term \emph{approximate} for $k$-NN search algorithms which may miss some of the $k$ nearest elements, and erroneously include some of the elements other than the truly nearest $k$.
% Najib: Same here with the math notation.
In other words, we call a $k$-NN search algorithm \emph{approximate} if there exists a query, dataset, distance function, or value of $k$ for which its set $H$ of returned points satisfies the condition $H \triangle S_q \neq \emptyset$, where $\triangle$ denotes a symmetric difference.
% Najib: When we refer to an \emph{exact} $k$-NN algorithm, we mean an algorithm which exhibits perfect recall when compared to a na\"{i}ve linear search.
When we refer to an \emph{exact} $k$-NN algorithm, we describe an algorithm which, no matter the query, dataset, distance function, nor value of $k$, returns precisely the $k$ nearest neighbors to that query;
no closer points are elided and no further points are included.

% Najib: Remove/replace the word ``pervasive'' here. It has a negative connotation.
$k$-nearest neighbors ($k$-NN) search is one of the most pervasive classification and recommendation methods in use~\cite{fix1952discriminatory, cover1967nearest}. 
Na\"{i}ve implementations of $k$-NN, whose time complexity is linear in the dataset's cardinality, prove prohibitively slow for large datasets. % Najib: because the dataset's cardinality is growing exponentially.
% Najib: While algorithms for fast $k$-NN on large datasets DO exist, most of them are approximate and do not exploit the geometric and topological structure inherent in SUCH datasets.
While algorithms for fast $k$-NN on large datasets exist, most of them are approximate and do not exploit the geometric and topological structure inherent in large datasets.
% Najib: Further, such algorithms are often approximate and while approximate search MAY be sufficient for some applications, the need for efficient AND \emph{exact} search remains.
Approximate search is sufficient for some applications, but the need for efficient, \emph{exact} search remains.
For example, for a majority voting classifier, approximate $k$-NN may agree with exact $k$-NN for large values of $k$, but may be sensitive to local perturbations with smaller values of $k$.
This is especially true when classes are not well-separated. % TODO: Cite something which says this.
Further, there is some evidence that distance functions which are non-metrics, such as cosine distance, perform poorly for $k$-NN in biomedical settings~\cite{hu2016distance};
this may suggest that approximate $k$-NN (which may have imperfect precision and recall) could exhibit suboptimal accuracy.

@@

This paper focuses on four novel algorithms for exact $k$-NN search.
We also present some improvements to the clustering and $\rho$-nearest neighbors search algorithms in CHESS, as well as improved genericity across distance functions.
% Najib: Cite HNSW here. Also, add ANNOY.
Finally, we provide a comparison of CAKES to the state-of-the-art in similarity search, FAISS~\cite{johnson2019billion} and HNSW, on several datasets from the ANN benchmarks suite~\cite{aumuller2020ann}.

### 2023-12-13 sections/2-methods.tex
some edits

In this manuscript, we are primarily concerned with $k$-nearest neighbors search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation (e.g., the genome of 
an organism, the neural embedding of a single image, or any entity on which we can define a \emph{distance function}).

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number.
We require that the distance function be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
In addition to these constraints, if the distance function obeys the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
Choice of distance function varies by type of data.
For example, with electromagnetic spectra, one could both Euclidean (L2) and Cosine distances.
With biological or string data, Hamming and Levenshtein distances are more appropriate.

@@

Figures 2 and 3 in~\cite{ishaq2019clustered} illustrate the low fractal dimensions of two datasets used with $\rho$-NN search in CHESS. 
We observe that for both APOGEE (Figure 2 in~\cite{ishaq2019clustered}) and GreenGenes (Figure 3 in~\cite{ishaq2019clustered}), other than the most extreme 10\% of clusters, virtually all clusters have a local fractal dimension of less than 2. 
This suggests that APOGEE and GreenGenes are good datasets for use with entropy-scaling search algorithms like CHESS and CAKES.

@@

We define a \emph{cluster} as a set of points with a \emph{center}, a \emph{radius}, and an approximation of the \emph{local fractal dimension}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance from any point in the cluster to the \emph{center}.
We estimate \emph{local fractal dimension} at the scale of the cluster radius and half the cluster radius.
Each cluster (unless it is a leaf cluster) has two child clusters in much the same way that a node in a binary tree has two child nodes.
We define the \emph{metric entropy} $\mathcal{N}_{\hat{r_c}}(X)$ of a data set $X$ under a hierarchical clustering scheme as a refinement of~\cite{yu2015entropy}, where metric entropy for a given cluster radius $r_c$ was the number of clusters of radius $r_c$ needed to cover all data.
Here, we use a hierarchical, divisive clustering approach, but with early stopping criteria; clusters which satisfy a user-specified stopping criterion (e.g. a specified cardinality or radius) are not further split.
Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of leaf clusters.

We start by building a divisive hierarchical clustering of the dataset with CLAM, using a similar recursive procedure as outlined in CHESS, but with the following improvements:
better selection of poles for partitioning and depth-first dataset reordering (see Section ~\ref{subsubsec:methods:clustering:dataset-depth-first-reordering}).

@@

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:sharding-and-auto-tuning}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.png}
    \caption{}
    \label{fig:methods:deltas}
\end{figure}

@@

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which could contain points in the query ball.

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-1.pdf}
    \caption{
        Triangle formed by query $q$, and poles $\ell$ and $r$. Query ball of radius $\rho$ centered at $q$. 
    }
    \label{fig:methods:overlapping-children-1}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-3.pdf}
    \caption{
        $m$ denotes the midpoint of $\overline{\ell r}$. $q'$ is the projection of $q$ onto $\overline{\ell r}$. $d$ is the distance from $q$ to the perpendiculat bisector of $\overline{\ell r}$. If $d \leq \rho$, both the left child and right child clusters could contain points in the query ball. Otherwise, only the right child cluster could contain such points. 
    }
    \label{fig:methods:overlapping-children-3}
\end{figure}

@@

We can do a quick check using values computed during partitioning (see Section~\ref{subsubsec:methods:cluster-partitioning}) to determine whether $d \leq \rho$. 
We note that $d = ||\overline{mq'}|| = ||\overline{mr \vphantom {q'}}||-||\overline{q'r}||$. Since $m$ is the midpoint of $\overline{\ell r}$, we have that $||\overline{mr}|| = \tfrac{||\overline{\ell r}||}{2}$, so \begin{equation} d = \tfrac{||\overline{\ell r}||}{2} - ||\overline{q'r}||.
\label{eq:methods:d-expression}
\end{equation}

@@

\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE{$\rho$-NN($cluster.left, query, r$)}
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
\end{algorithm}

@@

where $\hat{r_c}$ is the \textit{mean} cluster radius of leaf clusters, $\mathcal{N}_{\hat{r_c}}(X)$ is the metric entropy at that radius, $B_X(q,r)$ is a ball of data points around the query $q$ at search radius $r$, and $d$ is the fractal dimension around the query in a region of radius $r$.

>>>>

In this manuscript, we are primarily concerned with $k$-nearest neighbors search in a finite-dimensional space.
Given a dataset $\textbf{X} = \{x_1 \dots x_n\}$, we define a \emph{point} or \emph{datum} $x_i \in \textbf{X}$ as a singular observation, for example, the genome of an organism, the neural-network embedding of an image, etc.

We define a \emph{distance function} $f : \textbf{X} \times \textbf{X} \mapsto \mathbb{R}^+ \ \cup \ \{0\}$ as a function which, 
given two points $x, y \in \textbf{X}$, deterministically returns a non-negative real number.
We require that the distance function be symmetric (i.e., $f(x, y) = f(y, x)$ for all $x, y \in \textbf{X}$) and that the distance between two points $x$ and $y$ be zero if and only if $x = y$. 
In addition to these constraints, if the distance function obeys the triangle inequality, i.e., it is a \emph{distance metric}, then we guarantee that CAKES has perfect recall. 
The choice of distance function varies by dataset.
For example, with neural-network embeddings, one could use Euclidean (L2) distance or cosine distance.
With biological or string data, Levenshtein or Hamming distances are more appropriate.
With time-series data, one could use dynamic time warping (DTW) distance.

@@

% Najib: Add our own plots of local fractal dimension instead of referencing figures from the CHESS paper. Then, rewrite this paragraph.
Figures 2 and 3 in~\cite{ishaq2019clustered} illustrate the low fractal dimensions of two datasets used with $\rho$-NN search in CHESS. 
We observe that for both APOGEE (Figure 2 in~\cite{ishaq2019clustered}) and GreenGenes (Figure 3 in~\cite{ishaq2019clustered}), other than the most extreme 10\% of clusters, virtually all clusters have a local fractal dimension of less than 2. 
This suggests that APOGEE and GreenGenes are good datasets for use with entropy-scaling search algorithms like CHESS and CAKES.

@@

Najib We define a \emph{cluster} as a set of points with a \emph{center} and a \emph{radius}.
The \emph{center} is the geometric median of a sample of points in the \emph{cluster}, and so it is a real data point.
The \emph{radius} is the maximum distance of any point in the cluster to the \emph{center}.
We estimate \emph{local fractal dimension} at the scale of the cluster radius and half the cluster radius.
Each non-leaf cluster has two child clusters in much the same way that a node in a binary tree has two child nodes.
% Najib: Add an equation for local fractal dimension with the simplification from the previous sentence.

% Najib: Rewrite this paragraph. Old paper defined it as ... but we define it for the hierarchical clustering scheme as ...
We define the \emph{metric entropy} $\mathcal{N}_{\hat{r_c}}(X)$ of a data set $X$ under a hierarchical clustering scheme as a refinement of~\cite{yu2015entropy}, where metric entropy for a given cluster radius $r_c$ was the number of clusters of radius $r_c$ needed to cover all data.
Here, we use a hierarchical, divisive clustering approach, but with early stopping criteria; clusters which satisfy a user-specified stopping criterion (e.g. a specified cardinality or radius) are not further split.
Since we frame the asymptotic complexity of $\rho$-NN search in terms of the number of leaf clusters, the \emph{metric entropy} is best thought of in terms of the number of leaf clusters.

We start by building a divisive hierarchical clustering of the dataset using CLAM.
The procedure is similar to what was outlined in CHESS, but with the following improvements:
better selection of poles for partitioning and depth-first dataset reordering (see Section ~\ref{subsubsec:methods:clustering:dataset-depth-first-reordering}).

@@

\subsection {Sharding and Auto-Tuning}
\label{subsec:methods:sharding-and-auto-tuning}
% Najib: Remove Sharding from this subsection. We do not show sharding in any of our benchmarks. All benchmarks are done with datasets that fit in memory (on giant machines, but still).

@@

% Najib: For this figure, I want the red, green and blue lines/braces to be thinner. Perhaps 1/2 to 2/3 as thick.
% Najib: Also, add a caption.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.png}
    \caption{}
    \label{fig:methods:deltas}
\end{figure}

@@

We conduct $\rho$-nearest neighbors search as described in~\cite{ishaq2019clustered}, but with the following improvement:
when a cluster overlaps with the query ball, instead of always proceeding with both of its children, we proceed only with those children which could contain points in the query ball. 
% Najib: This is laid out in Algorithm 2. We should reference that algorithm here.

% Najib: Merge Figures 2 and 3. There's too much redundancy between them.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-1.pdf}
    \caption{
        Triangle formed by query $q$, and poles $\ell$ and $r$. Query ball of radius $\rho$ centered at $q$. 
    }
    \label{fig:methods:overlapping-children-1}
\end{figure}

@@

% Najib: Visually, ``m'' is not centered on the line segment. It's a little bit to the right.
\begin{figure}[ht!]
    \centering
    \includegraphics[scale=0.5]{images/geometry/overlapping-children-3.pdf}
    \caption{
        $m$ denotes the midpoint of $\overline{\ell r}$. $q'$ is the projection of $q$ onto $\overline{\ell r}$. $d$ is the distance from $q$ to the perpendiculat bisector of $\overline{\ell r}$. If $d \leq \rho$, both the left child and right child clusters could contain points in the query ball. Otherwise, only the right child cluster could contain such points. 
    }
    \label{fig:methods:overlapping-children-3}
\end{figure}

@@

% Najib: I don't like the double vertical bars for norm-notion. I think it's too messy. We should just overload the notation of the single horizontal bar across the top.
% Najib: Remove the equation numbers from this paragraph. They're not necessary.
We can do a quick check using values computed during partitioning (see Section~\ref{subsubsec:methods:cluster-partitioning}) to determine whether $d \leq \rho$. 
We note that $d = ||\overline{mq'}|| = ||\overline{mr \vphantom {q'}}||-||\overline{q'r}||$. Since $m$ is the midpoint of $\overline{\ell r}$, we have that $||\overline{mr}|| = \tfrac{||\overline{\ell r}||}{2}$, so \begin{equation} d = \tfrac{||\overline{\ell r}||}{2} - ||\overline{q'r}||.
\label{eq:methods:d-expression}
\end{equation}

@@

% Najib: Move the cluster.isLeaf check to the beginning of the algorithm because it's the base case.
\begin{algorithm} 
    \caption{$\rho$-NN(\emph{cluster, query, r})} 
    \label{alg:methods:rnn-search} 
    \begin{algorithmic}
        \REQUIRE $r \geq 0$
        \REQUIRE $cluster \neq \emptyset$
        \STATE $results \Leftarrow \emptyset$
        \IF{$cluster.left \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.left.\delta$ $\leq$ $r$ + $cluster.left.radius$}
                \STATE{$\rho$-NN($cluster.left, query, r$)}
            \ENDIF
        \ENDIF
        \IF{$cluster.right \cap B_X(query, r) \neq \emptyset$}
            \IF{$cluster.right.\delta$ $\leq$ $r$ + $cluster.right.radius$} 
                \STATE $\rho$-NN($cluster.right, query, r$)
            \ENDIF
        \ENDIF
        \IF{$cluster.isLeaf$}
            \FOR{$p \in cluster.points$}
                \IF{$p.\delta \leq r$}
                    \STATE $results.push((p, r))$
                \ENDIF
            \ENDFOR
        \ENDIF
        \STATE Return $results$
    \end{algorithmic}
\end{algorithm}

@@

where $\hat{r_c}$ is the \textit{mean} cluster radius of leaf clusters, $\mathcal{N}_{\hat{r_c}}(X)$ is the metric entropy at that radius, $B_X(q,r)$ is a ball of data points around the query $q$ at search radius $r$, and $d$ is the local fractal dimension around the query in a region of radius $r$.

### 2023-12-13 sections/4-results.tex
some edits

Figures~\ref{fig:results:fashion-mnist-scaling}, ~\ref{fig:results:glove-25-scaling}, and ~\ref{fig:results:sift-scaling} show the scaling behavior of CAKES on augmented versions of the following ann-benchmark datasets: fashion-mnist under Euclidean distance, deep-image under cosine distance, and glove-25 under cosine distance. 
Figure~\ref{fig:results:random-scaling} shows the scaling behavior of CAKES on a completely randomly generated dataset which does not obey any manifold structure. 
The the horizontal axes in each figure denote the multiplier by which we increased the cardinality of the original dataset with synthetic points (see Section \ref{subsec:methods:synthetic-data}). 
The vertical axes denote throughput in queries per second. Both axes are on a logarithmic scale. In this section, we discuss results only for $k = 10$. 
To see results with other values of $k$, refer to our Supplement section.

In Figure~\ref{fig:results:fashion-mnist-scaling}, which shows results for fashion-mnist, we observe that as the augmentation multiplier increases, each of CAKES's four algorithms exhibits significantly higher throughput than na\"{i}ve linear search, plotted in orange. 
Amongst them, GreedySieve, plotted in blue, consistently performs best, exhibiting sublinear time performance for nearly all cardinality multipliers. 
We see that for multipliers around $10^3$, Greedy Sieve exhibits throughput which is nearly three orders of magnitude higher than that of linear search.

With glove-25 (Figure~\ref{fig:results:glove-25-scaling}), we observe that each of our four algorithms outperforms linear search at high cardinality multipliers. 
On this dataset, Repeated $\rho$-NN, plotted in green, is consistently the best performing algorithm. {\color{red} Revive lfd violin plots to add explanation for this.}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/result_plots/fashion-mnist_10_scaling.png}
    \caption{
        Fashion-mnist Scaling
    }
    \label{fig:results:fashion-mnist-scaling}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/result_plots/glove-25_10_scaling.png}
    \caption{
        Glove-25 Scaling
    }
    \label{fig:results:glove-25-scaling}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/result_plots/sift_10_scaling.png}
    \caption{
        Sift Scaling
    }
    \label{fig:results:sift-scaling}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{images/result_plots/random_placeholder_10_scaling.png}
    \caption{
        Random Scaling
    }
    \label{fig:results:random-scaling}
\end{figure}

>>>>

Figures~\ref{fig:plots:fashion-mnist-scaling}A, ~\ref{fig:plots:glove-25-scaling}B, and ~\ref{fig:plots:sift-scaling}C show the scaling behavior of CAKES on augmented versions of the following ann-benchmark datasets: fashion-mnist under Euclidean distance, glove-25 under cosine distance, and sift under Euclidean distance. 
Figure~\ref{fig:plots:random-scaling}D shows the scaling behavior of CAKES on a completely randomly generated dataset with same cardinality and dimensionality as sift (1 million points in 128 dimensions).
The horizontal axis in each figure denotes the multiplier by which we increased the cardinality of the original dataset with synthetic points (see Section \ref{subsec:methods:synthetic-data}). 
The vertical axis denotes throughput in queries per second. Both axes are on a logarithmic scale. In this section, we discuss results only for $k$-NN search with $k = 10$, but similar plots for $k=100$ can be found in our Supplement section. In these plots, we label the recall at each scale for algorithms which do not exhibit perfect recall; if no such labels are given for an algorithm, then that algorithm exhibits perfect recall at all scales.

Tables \ref{table:results:ann-fashion}, \ref{table:results:ann-glove-25}, \ref{table:results:ann-sift}, and \ref{table:results:ann-random} show the throughput and recall of CAKES's algorithms at each cardinality multiplier (i.e., scale) for the fashion-mnist, glove-25, sift, and random datasets respectively. When reporting recall, we use 1.000* to denote that the recall is imperfect, but rounds to 1.000 when we consider
only three decimal places.

In Figure~\ref{fig:results:fashion-mnist-scaling}, which shows results for fashion-mnist, we observe that as the augmentation multiplier increases, the CAKES algorithms (GreedySieve in blue, RepeatedRnn in green, and Sieve in red) exhibit higher throughput than na\"{i}ve linear search, plotted in orange. We also observe that CAKES outperforms state-of-the-art similarity search algorithm FAISS-Flat (plotted in brown) for all cardinality multipliers and outperforms FAISS-IVF (plotted in pink) for all multipliers greater than $10^{5.5}$. Though state-of-the-art algorithms HNSW (plotted in gray), and Annoy (plotted in purple) have higher throughput than CAKES for all multipliers, we note that CAKES exhibits perfect recall for all multipliers, while hnsw and annoy exhibit much lower recall, as shown in Table~\ref{table:results:ann-fashion}. For example, at a cardinality multiplier as low as eight, hnsw and annoy have recall of 0.525 and 0.857 respectively, while CAKES exhibits perfect recall.
Amongst the CAKES algorithms, GreedySieve consistently performs best on this dataset, exhibiting sublinear time performance for nearly all cardinality multipliers.

With glove-25 (Figure~\ref{fig:results:glove-25-scaling}), we observe that each of CAKES' algorithms outperforms linear search at high cardinality multipliers. Similar to the results for fashion-mnist, we observe that CAKES outperforms FAISS-Flat for all cardinality multipliers and outperforms FAISS-IVF for all multipliers greater than $10^{7}$. We also see that while hnsw and annoy have higher throughput than CAKES for all multipliers, CAKES exhibits perfect recall for all multipliers, while hnsw and annoy exhibit much lower recall, as shown in Table~\ref{table:results:ann-glove-25}. For example, at a cardinality multiplier as low as eight, hnsw and annoy have recall of 0.294 and 0.857 respectively, while CAKES exhibits perfect recall.
On this dataset, Repeated $\rho$-NN, plotted in green, is consistently the best-performing algorithm. {\color{red} Revive lfd violin plots to add explanation for this.}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{plots/fashion-mnist-knn-10.png}
    \caption{
        Algorithm performance by cardinality multiplier on fashion-mnist for $k=10$.
    }
    \label{fig:results:fashion-mnist-scaling}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{plots/glove-25-knn-10.png}
    \caption{
        Algorithm performance by cardinality multiplier on glove-25 for $k=10$.  }
    \label{fig:results:glove-25-scaling}
\end{figure}

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{plots/sift-knn-10.png}
    \caption{
        Algorithm performance by cardinality multiplier on sift for $k=10$.    }
    \label{fig:results:sift-scaling}
\end{figure}

@@

\begin{figure}[ht!]
    \centering
    \includegraphics[width=3.4in]{plots/random-1000000-128-knn-10.png}
    \caption{
        Algorithm performance by cardinality multiplier on a random dataset for $k=10$.
    }
    \label{fig:results:random-scaling}
\end{figure}
